---
title: "Portfolio Project"
author: "Quinn Lynas"
warning: false
echo: false
embed-resources: true
---

**Data Description:** I sourced my data from kaggle. The website is here: https://www.kaggle.com/datasets/rohitsahoo/sales-forecasting/data. My data has information on sales of a large retailer from 2015 through 2019. This includes customer information and location, information on what the customer purchased, and how much was spent on the purchase. I want answer the following questions:

1.  What type of products were the most purchased between 2015 and 2019?

2.  What state had the most products purchased?

3.  How does the market in California look and has it changed?

4.  How are sales distributed between different sub-categories?

**Data Visualization:** I want some sort of time series visualizations so that we can see changes over time. I am guessing a line graph or scatter plot will work well for this, with the date on the x-axis. For my second research question a bar chart will work. For my third research question I will have to visualize using a graph that shows proportion well. I wanted to do a sankey/alluvial plot for this one. Missing data shouldn't be an issue.

**Data Cleaning:** The data set I am looking at has 9800 rows of information. I will need to fix the zip codes, some of them are missing and some only have 4 digits. I will need to manually fix this. After the zips are fixed, I will match the zip codes with data from the "zipdata" library, which has longitude and latitude information for each zip. I will also be looking at state and overall U.S. longitude and latitude coordinates so that I can answer question 2.

### Data Cleaning

The data set required some fixing with the zip codes. Some of the easterly states who's zip starts with zero mistakenly read in as only 4 digit zips. I had to put a zero at the front of every zip with that problem. On top of that, one of the cities had missing data for zip code, so I manually coded in the city's zip. I used a package called "zipcode" which contains a data set with longitude and latitude coordinates for each zip code in the United States. I joined this data set with the original so that each observation had coordinates. I also used the "lubridate" package to format the shipping dates better.

```{r}
#| label: Data cleaning
#| include: false
library(tidyverse)
library(ggplot2)
library(ggsankey)
library(ggalluvial)
library(zipcode)
library(lubridate)
library(sf)
library(mapdata)
library(maps)
data(zipcode) #get zip code data 
usa <- map_data("usa") #data for US map
state = map_data("state") #data for US states
sales <- read_csv("data/train.csv", col_types = c(`Postal Code` = "c")) #read in sales data
naniar::gg_miss_var(sales) #check for missing values in the data
sales <- janitor::clean_names(sales) #clean column names in the data

sales |>
  filter(is.na(postal_code)) #check where the zip code is na

burlington_zip = "05405" #set zip code for burlington which is the only city with NA zips

sales <- sales |>
  mutate(postal_code = case_when(
    is.na(postal_code) ~ burlington_zip,
    .default = postal_code
  )) #input postal code for missing values



sales |>
  filter(str_width(postal_code) == 4) #check to make sure zip codes are all 5 digits

sales <-  sales |>
  mutate(postal_code = case_when(
    str_width(postal_code) == 4 ~ paste0("0", postal_code),
    .default = postal_code
  )) #input 0 in front of zip codes that are length 4

sales <- sales |>
  mutate(order_date = dmy(order_date), ship_date = dmy(ship_date)) |>
  rename(zip = postal_code) #use lubridate to format dates, rename postal code so that we can join two datasets

zipcode <- zipcode |>
  select(zip, latitude, longitude) #get rid of unnecessary columns in zipcode data

sales <- left_join(sales, zipcode, by = join_by(zip)) #join sales and zipcode data by zip


```

### Visualizations 1 and 2

*Research Question 1:* What type of products were the most sold between 2015 and 2019?

The first two visualizations aim to answer my first research question, seen above. The visualization on the left shows the most sold sub-categories in order from least to most sold. You can see the two most sold products are phones and chairs. The next closest is storage, but the difference between them is quite large. The visualization on the right shows the most sold categories, in order from least to most sold. It is evident that technology sells the most.

```{r}
#| label: Question 1 Visualization
#| layout-ncol: 2
sales |>
  group_by(sub_category) |>
  summarise(n_sales = sum(sales)) |>
  ggplot(aes(x = n_sales / 1000, y = fct_reorder(sub_category, n_sales), fill = fct_reorder(sub_category, n_sales))) + 
  geom_col() + 
  labs(x = "Total Sales (in thousands)", y = "Sub-category", title = "Most sold products for a global superstore", caption = "Phones are most sold sub-category", alt = "Horizontal bar chart depicting total sales for different sub-categories in a superstore. Each horizontal bar represents a sub-category, which are sorted from least to most sales. Phones have the highest sales among all sub-categories. The x-axis uses a per 1000 scale.") +
  scale_x_continuous() + 
  theme_minimal() +
  scale_fill_viridis_d(option = "magma", direction = -1, begin = 0.2, end = 0.9) + 
  guides(fill = "none") 

sales |>
  group_by(category) |>
  summarise(n_sales = sum(sales)) |>
  ggplot(aes(y = n_sales / 1000, x = fct_reorder(category, n_sales), fill = fct_reorder(category, n_sales))) + 
  geom_col() +
  labs(y = "Total Sales (in thousands)", x = "Category", title = "Most sold products for a global superstore", caption = "Technology is most sold category by a wide margin", alt = "Vertical bar chart showing total sales by product category in a global superstore. Bars are sorted by sales. Technology has the highest sales, then Furniture and Office Supplies. The y-axis uses a per 1000 scale.") +
  theme_minimal() +
  scale_fill_viridis_d(option = "magma", direction = -1, begin = 0.2, end = 0.9) + 
  guides(fill = "none") 
```

### Visualizations 3, 4, 5, and 6

*Research Question 2:* What state has the most products purchased?

The next four visualizations aim to answer the above research question. The first visualization answers the research question, as you can clearly see California has the most total sales. This visualization and the next only show the states which are top 10 for total sales. The second visualization shows a map of the U.S. and the sales between each of the states. You can see there is a very high density for sales in California, and also states in the Northeast and Midwest. Since I found California had the most sales, I wanted to examine the sales within California to see if they were different than the country's sales. The order is relatively similar, phones and chairs hold the top two spots respectively. After examining the sub-categories of products, I decided to look into the hot-spots for sales in the state. My visual revealed most of the sales coming from the bay area, Los Angeles, and San Diego. Few sales came from the central valley.

```{r}
#| label: Question 2 Visualizations
#| layout-ncol: 2
top10_states <- sales |>
  group_by(state) |>
  summarise(state_sales = sum(sales)) |>
  arrange(desc(state_sales)) |>
  slice_head(n = 10)


top10_states |>
  ggplot(aes(x = state_sales / 1000, y = fct_reorder(state, state_sales), fill = fct_reorder(state, state_sales))) + 
  geom_col() + 
  labs(y = "", x = "Total Sales (in thousands)", title = "Top 10 Ranked States for Total Sales", caption = "California has the highest total sales", alt = "Horizontal bar chart showing total sales for each state that ranks top 10 in sales. Bars are sorted by total sales. California has the most total sales, followed by New York and others. The x-axis is in thousands of sales.") + 
  theme_minimal() + 
  guides(fill = "none") + 
  scale_fill_viridis_d(option = "magma", direction = -1, begin = 0.2, end = 0.9)


ggplot() +
  geom_polygon(data = state, aes(x = long, y = lat, group = group), fill = "grey90",  color = "black") + 
  geom_jitter(data = sales |>
                filter(state %in% top10_states$state), 
                aes(x = longitude, y = latitude, color = sales, size = sales), width = 0.5, height = 0.5) + 
  scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.9, trans = "sqrt") + 
  theme_void() +
  guides(color = "none", size = "none") +
  labs( x = "", y = "", title = "       States which are top 10 in Sales in the U.S.", caption = "Most sales in the country come from California,\nnortheast, midwest, Texas, and Florida", alt = "Map of the United States with points marked for each sale in the top 10 states. Location is marked for each point, and the size of each point is determined by how many sales there are. Color is transformed with a square root scale so that larger values don't impact the scale too much.") +
  theme(axis.text = element_blank(), legend.position = "bottom", aspect.ratio = 0.5) 

sales |>
  group_by(sub_category) |>
  filter(state == "California") |>
  summarise(n_sales = sum(sales)) |>
  ggplot(aes(y = fct_reorder(sub_category, n_sales), x = n_sales, fill = fct_reorder(sub_category, n_sales))) + 
  geom_col() +
  theme_minimal() +
  labs(x = "Total Sales", y = "", caption = "Phones and chairs rank 1st and 2nd for product\nsub-categories in CA, similar to the whole country's pattern", alt = "Horizontal bar chart showing total sales for each sub-category in California. Bars are sorted by total sales. Phones and chairs are the most sold products, similarly to the whole country's pattern.") + 
  scale_fill_viridis_d(option = "magma", direction = -1, begin = 0.2, end = 0.9) + 
  guides(fill = "none")

ggplot() + 
  geom_polygon(data = state |> 
                 filter(region == "california"), 
                 aes(x = long, y = lat, group = group), fill = "grey90", color = "black") + 
  geom_jitter(data = sales |>
               filter(state == "California"),
               aes(x = longitude, y = latitude, color = sales), alpha = 0.4, width = 0.1, height = 0.1) +
  scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8, trans = "log10", breaks = c(1, 10, 100, 1000)) +
  theme_void() + 
  theme(legend.position = "right", aspect.ratio = 1) + 
  labs(color = "Sales", caption = "Bay area and southern California account for most of the sales in the state.", alt = "Map of the State of California, with a point for each sale. Many of the sales are distributed near San Francisco and Los Angeles. The color of each point is mapped to the number of sales, with a brighter color meaning more sales. The points are jittered so overlapping points are more visible.")
```

### Visualization 7

*Research Question 3:* How does the market in California look and has it changed?

The 7th visualization below aims to answer my third research question. The line graph shows how product categories sales have changed from 2015 to 2019. Furniture and technology move similar to one another, while office supplies act more sporadically. Sales have increased for each category since 2015.

```{r}
#| label: Question 3 Visualization
df <- sales |>
  filter(state == "California") |>
  mutate(year = str_sub(order_date, start = 1, end = 4),
         year = as.numeric(year)) |>
  group_by(category, year) |>
  summarise(total_sales = sum(sales), .groups = "keep") |>
  ungroup()

ggplot(df, aes(x = year, color = category, y = total_sales)) + 
  geom_point() + 
  geom_line() + 
  scale_color_viridis_d(option = "magma", begin = 0.2, end = 0.9) +
  theme_classic() + 
  labs(color = "", y = "Total Sales", x = "Year", title = "Sales per Year for Different Product Categories in California", caption = "Office supplies see the strongest increase", alt = "Line graph with sales on the y-axis and year on the x-axis. Three separate lines represent different categories; furniture, office supplies, and technology. The office supplies see the largest increase in sales since the start date.")

```

### Visualization 8

*Research Question 4:* How are sales distributed between different sub-categories in California?

The 8th visualization aims to answer my fourth research question. The visual examines the different sale patterns within categories. Interestingly, office supplies has a lot of sales, but they are mostly low numbers. Furniture and technology have similar patterns, with technology having slightly higher sales but slightly less dense distribution.

```{r}
#| label: Question 4 Visualization
sales |>
  filter(state == "California") |>
  ggplot(aes(x = category, y = sales, color = fct_reorder(sub_category, sales))) + 
  ggbeeswarm::geom_beeswarm() + 
  scale_color_viridis_d(option = "magma") + 
  scale_y_continuous(trans = "log10") + 
  theme_classic() + 
  labs(y = "Sales", x = "Category", color = "Sub-category", caption = "Office supplies have low quantity sales, while furniture and tech have high quantity sales", alt = "Beeswarm graph with sales on the Y axis (log 10 transformed) and category on the x-axis. Three categories have points distributed for sales, each point represents sales of a sub-category. The color for subcategory is ordered by sales, subcategories with higher sales have brighter colors. While office supplies don't account for high sales, there are more low quantity sales than the other two categories.")
```
